---
title: "Exploring Mass Shootings in America"
author: "Benjamin Soltoff"
output: github_document
---

## Get the data

```{r, echo = TRUE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(knitr)        # load functions for formatting tables

# get data from rcfss package
# install latest version if not already installed
# devtools::install_github("uc-cfss/rcfss")
library(rcfss)

# load the data
data("mass_shootings")
mass_shootings
```

## Generate a data frame that summarizes the number of mass shootings per year. Print the data frame as a formatted `kable()` table.

```{r}
yearly_data <- mass_shootings %>%
count(year)

column_names <- c('Year', 'Number of mass shootings')


knitr::kable(
  yearly_data,
  format = "html",
  caption = "Mass shootings in the United States (1982-2019)",
  col.names = column_names,
)


```


## Generate a bar chart that identifies the number of mass shooters associated with each race category. The bars should be sorted from highest to lowest.

```{r}
data_without_na <- filter(.data = mass_shootings, !is.na(race))

ggplot(data = data_without_na) + 
  geom_bar(
    mapping = aes(x = forcats::fct_infreq(race)), 
    na.rm = TRUE
  ) + 
  labs(title = "Mass shootings in the United States (1982-2019)") +
  labs(x = "Race of perpetrator", y = "Number of incidents")


```

## Generate a boxplot visualizing the number of total victims, by type of location. Redraw the same plot, but remove the Las Vegas Strip massacre from the dataset.

```{r}
 
ggplot(data = mass_shootings, 
       mapping = aes(x = location_type, y = total_victims)) +
  geom_boxplot() +
  labs(title = "Mass shootings in the United States (1982-2019)") +
  labs(x = "Type of location of incident", y = "Total number of victims")

omitted_lvstrip <- filter(.data = mass_shootings, case!= "Las Vegas Strip massacre"	)

ggplot(data = omitted_lvstrip, 
       mapping = aes(x = location_type, y = total_victims)) +
  geom_boxplot() +
  labs(title = "Mass shootings in the United States (1982-2019)", subtitle = "Omitting Las Vegas Strip massacre") +
  labs(x = "Type of location of incident", y = "Total number of victims")
```

## How many white males with prior signs of mental illness initiated a mass shooting after 2000?

```{r}
x <- filter(.data = mass_shootings, male == TRUE & race == "White" & prior_mental_illness == "Yes" & year > 2000)
count(x)
```

**Answer** = 20 white males with prior signs of mental illness initiated a mass shooting after 2000.

## Which month of the year has the most mass shootings? Generate a bar chart sorted in chronological order to provide evidence of your answer.

```{r}

months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

ggplot(data = mass_shootings) +
  geom_bar( mapping = aes(x = month)) + 
   scale_x_discrete(limits = months) + 
  labs(title = "Mass shootings in the United States (1982-2019)") +
  labs(x = "Month", y = "Number of mass shootings")

```

**Answer**: From the above bar graph we can see that both February and June have the most mass shootings in the year. This is 12 mass shootings in both these months.

## How does the distribution of mass shooting fatalities differ between white and black shooters? What about white and latino shooters?

```{r}


fatality_race <- group_by(mass_shootings, fatalities, race)
races <- filter(.data = fatality_race, race == "White" | race == "Black" | race == "Latino") 
final <- summarize(races, 
          count_is = n()
          )

# ggplot(data = mass_shootings, mapping = aes(x = fatalities, y = ..count.., colour = race)) + 
  geom_line()

# ggplot( data = final, mapping = aes(x = fatalities, y = count_is)) +
  geom_histogram() +
  facet_grid(~race)

ggplot( data = mass_shootings, mapping = aes(x = fatalities)) +
  geom_histogram() +
  facet_grid(~race)

# ggplot(data = mass_shootings, mapping = aes(x = fatalities)) +
  stat_count(geom = 'line', aes(y = ..count.., color = races))


```

## Are mass shootings with shooters suffering from mental illness different from mass shootings with no signs of mental illness in the shooter? Assess the relationship between mental illness and total victims, mental illness and race, and the intersection of all three variables.

```{r}

# Plot 1: Relationship between mental illness and total victims
illness_without_na <- filter(.data = mass_shootings, !is.na(prior_mental_illness))

ggplot(illness_without_na, mapping = aes(x = prior_mental_illness, y = total_victims)) + 
  geom_boxplot() + 
  labs(title = "Mass shootings in the United States (1982-2019)") +
  labs(x = "Evidence of prior mental illness by perpetrator", y = "Number of total victims")

# Plot 2: Relationship between Race and Total Victimes
race_without_na <- filter(.data = mass_shootings, !is.na(race))

plot <- ggplot(data = race_without_na, 
       mapping = aes(x = race, y = total_victims)) +
  geom_boxplot() +
  labs(title = "Mass shootings in the United States (1982-2019)") +
  labs(x = "Race of perpetrator", y = "Total number of victims")

plot 

# Plot 3 ISSUE
plot + ylim(0,100)
  # scale_y_continuous(breaks=c(seq(3,10,7), seq(10, 30, 20), seq(30,100,70), seq(100,300,200)))

# Plot 4: Omitting Las Vegas Strip massacre

omitted_lvstrip <- filter(.data = mass_shootings, case!= "Las Vegas Strip massacre"	)

ggplot(data = omitted_lvstrip, 
       mapping = aes(x = race, y = total_victims)) +
  geom_boxplot() +
  labs(title = "Mass shootings in the United States (1982-2019)", subtitle = "Omitting Las Vegas Strip Massacre (2017") +
  labs(x = "Race of perpetrator", y = "Total number of victims")

# Plot 5: Intersection of Race, Total Victims and Prior Mental Illness

clean_data <- filter(.data = mass_shootings, race!= "Other" & !is.na(race) & !is.na(prior_mental_illness))
  
ggplot(data = clean_data, mapping = aes(x = total_victims, y = race, fill = prior_mental_illness)) +
  geom_boxplot() + 
  labs(title = "Mass shootings in the United States (1982-2019)") +
  labs(x = "Number of total victims", y = "Race of perpetrator") + 
  labs(fill = "Evidence of prior mental illness") + 
  theme(legend.position = "bottom")

# Plot 6: Another intersection of Race, Total Victims and Prior Mental Illness

clean_data <- filter(.data = mass_shootings, race!= "Other" & !is.na(race) & !is.na(prior_mental_illness))

ggplot(data = clean_data, mapping = aes(x = prior_mental_illness, y = total_victims, fill = race)) +
  geom_boxplot() + 
  labs(title = "Mass shootings in the United States (1982-2019)") +
  labs(x = "Evidence of prior mental illness", y = " Number of total victims") + 
  labs(fill = "Race of perpetrator") + 
  theme(legend.position = "bottom")


  

```
**Answer**: 
- From Plot 5, those with prior mental illesses have a longer box plots as well as longer whiskers when compared to those with no evidence of prior mental illness. This implies more variance in the total number of victims. Moreover, the outliers are also larger for those with prior mental illness. In contrast, the minimum value of total number of victims for those with prior mental illness is also smaller than those without. Moreover, the median values for both with and without illnesses is not very different. This leads to the conclusion that those with prior mental illnesses are more likely to be shooters, but also that their behaviour seems to be more sporadic and variable than those with no prior mental illness. This is also confirmed from Plot 1, where the medians are very close, but perpetrators with prior mental illnesses have higher maximum values, lower minimum values and larger variance.

- From Plot 6, having a mental illness creates perpetrators from all 5 races, whereas those with no prior mental illness belong to only 3 races. Moreover, having a mental illness leads to larger variance as mentioned above. 


## Session info

```{r, echo = TRUE}
devtools::session_info()
```


